\documentclass[../m2r-report.tex]{subfile}

\subsection{Finite Graphs}
\label{sec:1.1}

Whilst Ramsey's original paper \cite{Ramsey1930} is concerned with a problem of logic, Erd{\"o}s and 
Szekeres \cite{Erdos1935} reformulated his work in a graph theoretical sense. The idea of proving 
the existence of regular substructure within an apparently chaotic superstructure
is immediately applicable to graphs. The content of this section primarily follows 
that of Bollobás's book \textit{Modern Graph Theory} \cite{Bollobas1998}.

In the graph below, we partition the edges into two disjoint sets: red lines and blue 
lines. In turns out that no matter how we partition our edges, we will always get
red triangles or blue triangles. In this particular partition, we get two 
red triangles.

\begin{figure}[h]
\begin{center}
    \begin{tikzpicture}
        \draw [blue] (0,0) -- (2,0);
        \draw [blue] (2,0) -- (3,1.7321);
        \draw [blue] (3,1.7321) -- (2,3.4641);
        \draw [blue] (2,3.4641) -- (0,3.4641);
        \draw [blue] (0,3.4641) -- (-1,1.7321);
        \draw [blue] (-1,1.7321) -- (0,0);
        \draw [blue] (0,0) -- (2,3.4641);
        \draw [blue] (3,1.7321) -- (-1,1.7321);
        \draw [blue] (2,0) -- (0,3.4641);

        \draw [red] (0,0) -- (3,1.7321);
        \draw [red] (3,1.7321) -- (0,3.4641);
        \draw [red] (0,3.4641) -- (0,0);
        \draw [red] (2,0) -- (2,3.4641);
        \draw [red] (2,3.4641) -- (-1,1.7321);
        \draw [red] (-1,1.72,0) -- (2,0);

        \filldraw (0,0) circle (2pt);
        \filldraw (2,0) circle (2pt);
        \filldraw (3,1.7321) circle (2pt);
        \filldraw (2,3.4641) circle (2pt);
        \filldraw (0,3.4641) circle (2pt);
        \filldraw (-1,1.7321) circle (2pt);
    \end{tikzpicture}

\end{center}

\caption{2 red triangles within a graph of 6 vertices}
\end{figure}

Before continuing, we first need to state some notation and definitions \cite{Graham1990}.

For $n,k\in \Z^+$, the set of positive integers, and any non-empty set $X$, let
\begin{align*}%\label{eq:bracketDef}
[n] &= \{1, 2,\ldots, n\},\\
X^{(k)} &= \{Y:Y\subseteq X, |Y|=k\}.
\end{align*}

\begin{definition}[Graph]

    A \textit{graph G} is an ordered pair of disjoint countable sets $(V,E)$ such that $E$ 
    is a subset of $V^{(2)}$, the set of unordered pairs of $V$. We call $V$ the 
    set of vertices, and $E$ the set of edges. The \textit{order}
    of $G$ is the number of vertices, i.e $|V|$.

\end{definition}

\begin{definition}[Complete graph]

    We say a graph $G:=(V,E)$ is \textit{complete} if $E=V^{(2)}$. Intuitively, a graph 
    is complete if every pair of vertices has an edge connecting them. We denote the complete 
    graph of order $n$ by $K_n$.

\end{definition}

\begin{definition}[Subgraphs and Cliques]

    We say $G':=(V',E')$ is a \textit{subgraph} of $G:=(V,E)$ if $V'\subseteq V$, 
    and $E'\subseteq E$. We write $G'\subseteq G$. We say $G'$ is an \textit{induced} 
    subgraph of $G$ if $E'=E\cap V'^{(2)}$. We say an induced subgraph $G'$ is a \textit{clique} of $G$ if it is complete.

\end{definition}

Earlier we mentioned partitioning the edges into disjoint classes. We can think 
of these classes as being `colourings' of the edges. In an $r$-\textit{colouring},
we partition our edges into $r$ disjoint classes, called \textit{colours}. The 
following definition helps explain this:

\begin{definition}[Colouring of a set]

    Let $X$ be a non-empty set. For $r\in\Z^+$, an $r$-\textit{colouring} of $X$ is a 
    partition of $X$ into $r$ equivalence classes.
    
    Equivalently, a $r$-colouring of $X$ is the partition of $X$ induced by an 
    arbitrary map: $$\chi: X \to [r]$$ through the equivalence relation: 
    $$\mathcal{R}_{\chi}:= \{(x_{1},x_{2})\in X | \chi(x_{1})=\chi(x_{2}) \}$$
    Here $\chi$ is called the colouring function. Somewhat improperly, 
    we will identify a $r$-colouring of $X$ with its colouring function.

    In the context of graphs, an $r$-colouring of a graph $G=(V,E)$ is a map \cite{Graham1990}:

    $$\chi : E \to [r]$$

    We say a subgraph $G' = (V',E')$ of $G$ is \textit{monochromatic} if the restriction
    $\chi|_{E'}$ is the constant map.

\end{definition}

\begin{definition}[Ramsey number]

    For $s,t\in\Z^+$, we define the \textit{Ramsey number} of $s$ and $t$, $R(s,t)$, 
    to be the least positive integer such that any 2-colouring of $K_n$ contains
    as a clique a monochromatic $K_s$ or a monochromatic $K_t$, whenever 
    $n\geq R(s,t)$.

\end{definition}

We can now state the simplest form of Ramsey's theorem in a finite graph theoretical 
sense.

\begin{prop}\label{prop:ram}

    $R(s,t)$ is a well-defined function for all $s,t\in\Z^+$, and
    $$R(s,t)\leq R(s-1,t)+R(s,t-1).$$
\end{prop}

\textbf{Remark:} If $\chi$ is a 2-colouring of $K_n$, I will from now on refer 
to the edges $\chi^{-1}\{1\}$ as being `red',
and the edges $\chi^{-1}\{2\}$ as being `blue'.

\begin{proof}
    We proceed by double induction on $s$ and $t$. Note that $R(s,2) = R(2,s)=s$, 
    since a 2-colouring of $K_s$ either has a blue edge, or 
    every other edge is red. Hence $R(s,2),R(2,s),R(t,2),R(2,t)$ all exist. \cite{Bollobas1998}\\

    We now assume for an induction 
    that $R(s,t-1)$ and $R(s-1,t)$ both exist. Let $n=R(s-1,t)+R(s,t-1)$ and consider any
    2-colouring of the edges of $K_n$. We need to show that we have
    either a red $K_s$ or a blue $K_t$. So let $x$ be any vertex of
    $K_n$, and let:
    \begin{align*}
        T_1 &= \text{set of vertices connected to } x \text{ by red edges,} \\
        T_2 &= \text{set of vertices connected to } x \text{ by blue edges.}
    \end{align*}  

    Let $n_1 = |T_1|$ and $n_2 = |T_2|$. Then $n_1 + n_2 + 1 = R(s-1,t)+R(s,t-1)$.
    If $n_1 < R(s-1,t)$, then $n_2 \geq R(s,t-1)$, so $T_2$ contains a 
    red $K_s$, or a blue $K_{t-1}$. If the former holds then we are done, and if the latter holds then $K_{t-1}$ 
    forms a blue clique, $K_t$, with $\{x\}$ and the inequality is established.\\
    If instead $n_1\geq R(s-1,t)$, then $T_1$ contains either a red $K_{s-1}$, or 
    a blue $K_t$. If the latter holds then we are done, and if the former holds 
    then $K_{s-1}$ forms a red clique, $K_s$, with $\{x\}$.\cite{Greenwood1955}
\end{proof}

I now return to example presented earlier:

\begin{example}[$R(3,3)=6$]
\end{example}

\begin{proof}
    Notice that $R(3,3) \leq 2R(3,2) = 6$. To show that $R(3,3)>5$, we assume for 
    a contradiction that $R(3,3)=5$. Then for every 2-colouring of $K_5$, we should 
    be able to find a monochromatic $K_3$. However, if we consider the colouring 
    in which we colour the exterior edges of $K_5$ red, and the interior diagonal 
    edges blue, shown in the graph below, we see that there are no red triangles,
    nor blue triangles.

    \begin{figure}[h]
        \begin{center}
            \begin{tikzpicture}
                \draw [red] (0,0) -- (2,0);
                \draw [red] (2,0) -- (2.618,1.9022);
                \draw [red] (2.618,1.9022) -- (1,3.0776);
                \draw [red] (1,3.0776) -- (-0.618,1.9022);
                \draw [red] (-0.618,1.9022) -- (0,0);

                \draw [blue] (0,0) -- (2.618,1.9022);
                \draw [blue] (0,0) -- (1,3.0776);
                \draw [blue] (2,0) -- (1,3.0776);
                \draw [blue] (2,0) -- (-0.618,1.9022);
                \draw [blue] (2.618,1.9022) -- (-0.618,1.9022);
        
                \filldraw (0,0) circle (2pt);
                \filldraw (2,0) circle (2pt);
                \filldraw (2.618,1.9022) circle (2pt);
                \filldraw (1,3.0776) circle (2pt);
                \filldraw (-0.618,1.9022) circle (2pt);
            \end{tikzpicture}
        
        \end{center}
        
        \caption{A particular 2-colouring of $K_5$}
    \end{figure}
    We have our contradiction. Hence $R(3,3)=6$.\cite{Greenwood1955}
\end{proof}

\textbf{Remark}: In the same paper \cite{Greenwood1955}, Greenwood and Gleason also proved the 
values of the following Ramsey numbers: $R(3,4)=9$, $R(3,5)=14$ and $R(4,4)=18$. 
As of the date of writing, only small Ramsey numbers are actually known exactly. 
$R(5,5)$ is unknown, although in 1995 McKay and Radziszowski proved that its 
value must lie between 43 and 49 inclusive \cite{McKay1995}. Its actual value is unknown,  
and the exact values of Ramsey numbers are often computed using algorithms, and very 
powerful computers. Using their bounds, even for a simple $2$-colouring, a 
computer would have to check at least $2^{43}\approx 8.8$ trillion different colour 
configurations of $K_5$.\\

Greenwood and Gleason also showed that $R_3(3,3,3)=17$, i.e. every complete graph 
of order at least 17 has a monochromatic $K_3$, when coloured with 
three colours. In fact, \autoref{prop:ram} holds for any finite $r$-colouring.

\begin{theorem} \label{thr:ram2}

    Given $r\in\Z^+$ positive integers, $s_1, s_2,\ldots, s_r$, there exist a least
    positive integer, $R_r(s_1,\ldots,s_r)$, such that any r-colouring of $K_n$ with 
    $n\geq R_r(s_1,\ldots,s_r)$ contains, as a clique, a monochromatic $K_{s_i}$
    for some $i\in\{1,\ldots ,r\}$, coloured by the $i^{\text{th}}$ colour.\label{thr:genRamsey}

\end{theorem}

\begin{proof}

    We again proceed by induction on $r$. The base case for $R_2(s_1,s_2)$ is
    true by \autoref{prop:ram}. So we assume by induction that
    $R_{r-1}(s_1,\ldots ,s_{r-1})$ exists. Then in any $r$-colouring of $K_n$ we 
    group the first two colours into a new colour. So long as $n$ is large enough,
    either there is a $K_{s_i}$ coloured with the $i^{\text{th}}$ colour for
    some $i \in \{3,\ldots,r\}$ as a clique of $K_n$, or there is a clique $K_m$
    coloured with the new colour, with $m=R(s_1,s_2)$. 
    If the former holds, then we have our monochromatic clique. And if the latter holds, then 
    we have a monochromatic $K_{s_1}$ or a monochromatic $K_{s_2}$ as a clique of $K_m$ by \autoref{prop:ram}, 
    since $K_m$ is coloured by the first two colours of the original colouring.
    And so we are done by induction, since $K_m\subset K_n$.\cite{Bollobas1998}
\end{proof}

Whilst we have been discussing Ramsey's theorem in a strictly graph-theoretical sense,
the theory extends beyond that. If we are given a set $V$ (which we 
have been referring to as the set of vertices of a given graph), so far we have been 
looking at partitioning (colouring) the set $V^{(2)}$ (the edges) into $r$ classes. 
However, \autoref{thr:ram2} still holds when we partition $V^{(k)}$ into $r$ classes, 
for any $k,r\in\Z^+$.

\begin{theorem}[Ramsey (finite version)] \label{thr:hyper}

    Given $r\in\Z^+$ positive integers, $s_1, s_2,\ldots, s_r$, such that $1<k<\text{min}\{s_1,\ldots, s_r\}$, there exist a least
    positive integer, $R_r^{(k)}(s_1,\ldots,s_r)$, such that any $r$-colouring of $[n]^{(k)}$ 
    with $n\geq R_r^{(k)}(s_1,\ldots,s_r)$ contains, as a clique, a monochromatic 
    $[s_i]^{(k)}$ for some $i\in\{1,\ldots ,r\}$, coloured by the $i^{\text{th}}$ colour.  \cite{Ramsey1930}

\end{theorem}

\begin{proof}
    For simplicity, we will just consider the case $r=2$. The proof is very similar to the 
    one for \autoref{prop:ram}, and we will use triple induction on $s,t,k\in\Z^+$. 
    We will assume that both $R^{(k)}(s-1,t)$ and $R^{(k)}(s,t-1)$ exist, and 
    also $R^{(k-1)}(u,v)$ exists for all $u,v\in\Z^+$.\\

    Let $X$ be a set with $R^{(k-1)}(R^{(k)}(s-1,t),R^{(k)}(s,t-1))+1$ elements.
    We fix a 2-colouring $\chi:X^{(k)}\to [2]$ and an $x\in X$, and define 
    another 2-colouring $\chi':Y^{(k-1)}\to [2]$, where $Y=X\setminus\{x\}$, 
    given by $\chi'(\sigma) = \chi(\sigma\cup\{x\})$ for all $\sigma\in Y^{(k-1)}$. 
    By the definition of $R^{(k-1)}(u,v)$, we assume that $Y$ has a red (for $\chi'$) 
    subset $Z$ with $R^{(k)}(s-1,t)$ elements.\\

    Since $|Z|=R^{(k)}(s-1,t)$, then there is either a blue (for $\chi$) subset of $Z$ of size $t$, or  
    a red subset of size $s-1$. If the former holds, then we are done since $Z^{(k)}\subset X^{(k)}$. 
    If the latter holds, then we have a red subset of size $s-1$ which forms a red clique $K_s$ with $\{x\}$, since $\{x\}\cup\sigma$ 
    is red for all $\sigma\in Z^{(k-1)}$.\cite{Bollobas1998}
\end{proof}

\textbf{Remark}: We can use the same argument to prove that \autoref{thr:hyper} is in fact true for 
$r$-colourings. That is, $R_r^{(k)}(s_1,\ldots,s_r)$ exists. We now consider the case of infinite graphs.

\subsection{Infinite Graphs}
\label{sec:1.2}

The statement of Ramsey's theorem in the introduction
concerns partitioning a countably infinite set $\Gamma$ into $r$ disjoint 
subsets, where $r\in\Z^+$.

\begin{definition}[Infinite complete graph]

    We define the \textit{infinite complete graph}, $K_\infty$, as the ordered pair
    $K_\infty:=(\Gamma,\Gamma^{(2)})$.

\end{definition}

\begin{theorem}[Ramsey (infinite version)]\label{thr:ram}

    For any r-colouring of the edges of $K_\infty$, say
    $\chi :\Gamma^{(2)}\to\{c_1,\ldots, c_r\}$, there exists 
    within it a countably infinite monochromatic clique. In other words, there 
    exists $X \subset K_\infty$ such that $X$ is countably infinite, and $\chi|_{X^{(2)}}$ is the
    constant function.\cite{Narayana2017}

\end{theorem}

\textbf{Remark}: It should be noted that whilst \autoref{thr:ram} is referring to 
the edges of the infinite graph, $\Gamma^{(2)}$, Ramsey's original theorem states 
that the same result holds for any $r$-colouring of $\Gamma^{(k)}$, for 
any $k,r \in \Z^+$. That is, we can always find a countably infinite monochromatic 
subset $\Delta \subset \Gamma$.

\begin{proof}

    Fix an $r$-colouring of $K_\infty$ and $a_1 \in \Gamma$. Then by the pigeonhole principal, 
    there must exist a countably infinite set $S_1\subset\Gamma\setminus
    \{a_1\}$ such that all of the $a_1-S_1$ edges are of the same colour, say $c_1$. In other words, 
    $$S_1=\{x\in\Gamma\setminus\{a_1\}:\chi(\{a_1,x\})=c_1\}\text{ is countable.}$$

    Now again fix an arbitrary $a_2\in S_1$. Again, we find some countable
    $S_2\subset S_1$ such that all of the $a_2-S_2$
    are of the same colour, say $c_2$.\\

    If we repeat this inductively, we obtain a sequence $a_1,a_2,\ldots$ 
    and a sequence of colours $c_1,c_2,\ldots$ such that $\chi(\{a_i,a_j\})=
    c_i$, for $i<j$.\\

    Now by the pigeonhole principal again, since there are finitely many colours,
    there exists an countable subsequence $c_{i_1}, c_{i_2}, \ldots$ that is 
    constant, i.e. each term is equal to $c_i$. Then $\{a_{i_1},a_{i_2}, \ldots\}$ is an infinite
    monochromatic subset, coloured by $c_i$.\cite{Narayana2017}

\end{proof}

An interesting corollary of this result is the following key theorem from 
analysis:

\textbf{Corollary}: [Bolzano-Weierstrass Theorem]

Let $(x_i)_{i\in \Z_+}$ be a bounded sequence of real numbers. Then it has
a convergent subsequence.

\begin{proof}

    We define a 2-colouring of ${\Z^+}^{(2)}$ by calling $\{i,j\}$ \textit{red} 
    if $x_i <x_j$, and \textit{blue} if $x_i\geq x_j$.

    Then \autoref{thr:ram} gives us a countably infinite monochromatic 
    set, which in this case is a monotone subsequence of $(x_i)$. And since it 
    is bounded, by the monotone convergence theorem it must converge.\cite{Narayana2017}

\end{proof}

\subsection{Generalised Ramsey Numbers}
\label{sec:1.3}

The substructures that Ramsey numbers determine the existence of are the monochromatic 
cliques $K_s$ which we have so far exclusively discussed. This section is concerned with generalising 
the type of monochromatic substructure that we can assure the existence of, to an arbitrary 
graph $G$.

\begin{definition}[Generalised Ramsey number]
    Let $G$ be an arbitrary graph. We define the \textit{generalised 
    Ramsey number}, $r(G)$, to be the minimal $n\in\Z^+$ such that every 2-colouring of $K_n$
    contains a monochromatic $G$. \cite{Graham1990}

    Note that $R(s,s)=r(K_s)$, since $R(s,s)$ is the minimal $n$ such that every 
    2-colouring of $K_n$ contains as a clique a monochromatic $K_s$.

\end{definition}

Whilst the Ramsey theorems discussed so far, such as \autoref{thr:hyper} and 
\autoref{thr:ram}, do not easily extend to cover generalised Ramsey numbers, we 
give the following theorem as an elegant result in the emerging field of Graph 
Ramsey theory.

\begin{theorem}
    Let $nK_3$ be the graph consisting of $n\geq2$ disjoint triangles $K_3$. Then every 
    2-colouring of $K_r$ has a monochromatic $nK_3$ whenever $r\geq5n$. In other 
    words,\cite{Graham1990}
    $$r(nK_3)=5n\qquad \text{for  } n\geq2.$$
\end{theorem}

\begin{proof}
    We first show 
    that $r(nK_3)\geq5n$. Consider the 2-colouring of $K_{5n-1}$ in which we 
    form red triangles from any $3n-1$ vertices, blue triangles from the remaining $2n$ vertices, 
    and colour every other edge blue. Then the only monochromatic triangles here are 
    non-disjoint, so this colouring contains no monochromatic $nK_3$, and 
    so $r(nK_3)>5n-1$.\\

    We now show that $r(nK_3)\leq5n$ for $n\geq2$, by induction on $n$. The base 
    case $n=2$ requires detailed and unenlightening analysis, which we omit. Now fix 
    an arbitrary 2-colouring of $K_{5n}$, for $n\geq3$.
    We can find $\frac{5n-5}{3}$ disjoint monochromatic triangles by 
    simply selecting triangles from $K_{5n}$, until less than six vertices remain (since $R(3,3)=6$). Since 
    $\frac{5n-5}{3}\geq n$, we can assume that one of these triangles is monochromatic,
    say $\{1,2,3\}$ is red and $\{4,5,6\}$ is blue.\\

    We assume, by symmetry, that at least five of the edges 
    between $\{1,2,3\}$ and $\{4,5,6\}$ are blue. Then at least one of $\{1,2,3\}$, say 1, 
    is joined by blue edges to at least two of $\{4,5,6\}$, say 4 and 5, and 
    $\{1,2,3,4,5\}$ forms a bow-tie (by `bow-tie' we mean a pair of triangles of different colours that share a 
    single common vertex). When these five vertices are removed, the remaining $5n-5$ vertices of $K_{5n}$
    contain, by induction, a monochromatic $(n-1)K_3$. This, 
    together with a $K_3$ of the same colour from the bow-tie, forms a monochromatic $nK_3$.
    \cite{Graham1990}
\end{proof}

\textbf{Remark}: This proof can be generalised to find the generalised Ramsey 
numbers $r(sG,tG)$ for an arbitrary graph $G$ and any $s,t\in\Z^+$. For example, if $s\geq t\geq1$, then,
$$r(sK_2,tK_2)=2s+t-1,$$
where $sK_2$ is the graph consisting of $s$ disjoint edges. \cite{Bollobas1998}